---
title: "Reproducible Research Course Project 1"
output: html_document
---

#### Step 1. Loading and preprocessing the data

```{r}
act.df <- read.csv("activity.csv",header = TRUE, na.strings = "NA",stringsAsFactors = FALSE)
```

#### Step 2. What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day 

```{r}
# Clean up NAs and calculate sum of steps per day
act.sum.df <- aggregate(steps ~ date, data = act.df, sum,na.action = na.omit)

# Histogram of the total number of steps taken each day
hist(act.sum.df$steps, main="Histogram of total number of steps per day", xlab="Total number of steps per day")
```
2. Calculate and report the mean and median total number of steps taken per day

```{r}
# Mean of the total number of steps taken per day
mean(act.sum.df$steps)

# Median of the total number of steps taken per day
median(act.sum.df$steps)
```
*So the mean and median of the total number of steps taken per day are `r floor(mean(act.sum.df$steps))` and `r median(act.sum.df$steps)` respectively*

#### Step 3. What is the average daily activity pattern?
1. Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).


```{r}
# aggregate steps as interval to get average number of steps in an interval across all days
act.clean.df <- act.df[complete.cases(act.df),]
act.int.df <- aggregate(steps ~ interval, act.clean.df , mean)

# generate the line plot of the 5-minute interval (x-axis) and the average number of 
# steps taken, averaged across all days (y-axis)
plot(act.int.df$interval, act.int.df$steps, type='l', col=1, 
     main="Average number of steps averaged over all days", xlab="Interval", 
     ylab="Average number of steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
# find row id of maximum average number of steps in an interval
max.avg.steps.df<- which.max(act.int.df$steps)

# get the interval with maximum average number of steps in an interval
act.int.df [max.avg.steps.df, ]
```

### Step 4. Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs).

```{r}
# get rows with NA's
df.NA <- act.df[!complete.cases(act.df),]

# number of rows
nrow(df.NA)
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r}
# perform the imputation
for (i in 1:nrow(act.df)){
  if (is.na(act.df$steps[i])){
    interval_val <- act.df$interval[i]
    row_id <- which(act.int.df$interval == interval_val)
    steps_val <- act.int.df$steps[row_id]
    act.df$steps[i] <- steps_val
  }
}

# aggregate steps as per date to get total number of steps in a day
table_date_steps_imputed <- aggregate(steps ~ date, act.df, sum)

# create histogram of total number of steps in a day
hist(table_date_steps_imputed$steps,  main="(Imputed) Histogram of total number of steps per day", xlab="Total number of steps in a day")
```

